***De novo* clustering methods out-perform reference-based methods for assigning 16S rRNA gene sequences to operational taxonomic units**

### 2015-10-28 Schloss_Cluster_PeerJ_2015.Rmd
Edit manuscript

**Methods.** Our analysis implemented five *de novo* clustering algorithms including the single linkage, complete linkage, average linkage, abundance-based greedy clustering, distance-based greedy clustering and two reference-based methods including the open and closed-reference methods. By varying the number of sequences sampled from each of two previously published datasets we used the Matthew’s Correlation Coefficient (MCC) to assess the quality and stability of OTU assignments.

@@

**Discussion.** Our analysis calls into question the quality and stability of OTU assignments generated by the open and closed-reference methods as implemented in current version of QIIME. This study demonstrates that *de novo* methods are the most rigorous and that the quality of clustering assignments needs to be assessed for multiple methods to identify the optimal clustering method for the dataset.

@@

The ability to affordably generate millions of 16S rRNA gene sequences has allowed microbial ecologists to thoroughly characterize the microbial community composition of hundreds of samples. To simplify the complexity of these large datasets, it is helpful to cluster sequences into meaningful bins. These bins, commonly known as operational taxonomic units (OTUs), are  used to compare the biodiversity contained within and between different samples [@Schloss2011]. Such comparisons have enabled researchers to characterize the microbiota associated with the human body [e.g. @Huttenhower2012], soil [e.g. @Shade2013], aquatic ecosystems [e.g. @Gilbert2011], and numerous other environments. Within the field of microbial ecology, a convention has emerged where sequences are clustered into OTUs using a threshold of 97% similarity or a distance of 3%. One advantage of the OTU-based approach is that the definition of the bins is operational and can be changed to suit the needs of the particular project. However, with the dissemination of clustering methods within software such as mothur [@Schloss2009], QIIME [@Caporaso2010], and other tools [@Edgar2010; @Sun2009; @Mah2014; @Edgar2013; @Cai2011], it is important to understand how different clustering methods implement this conventional OTU threshold. Furthermore, it is necessary to understand how method choice affects the precision and accuracy of assigning sequences to OTUs. Broadly speaking, three approaches have been developed to assign sequences to OTUs.

The first approach has been referred to as phylotyping [@Schloss2011] or closed reference clustering [@NavasMolina2013]. This approach involves comparing sequences to a curated database and then clustering sequences into the same OTU that are similar to the same reference sequence. Reference-based clustering methods suffer when the reference does not adequately reflect the biodiversity of the community. If a large fraction of sequences are novel, then they cannot be assigned to an OTU. In addition, the reference sequences used in this application are selected because they are less than 97% similar to each other over the full length of the gene; however, it is known that the commonly used variable regions within the 16S rRNA gene do not evolve at the same rate as the full-length gene [@Schloss2011]. Thus, a sequence representing a fragment of the gene may be more than 97% similar to multiple reference sequences. Therefore, defining OTUs in the closed-reference approach is complicated because although two sequences might be 97% similar to the same reference sequence, they may only be 94% similar to each other. A subtle alternative to this approach is to use a classifier to assign a taxonomy to each sequence so that sequences can be clustered at a desired level within the Linnean taxonomic hierarchy [@Schloss2011]. The strength of the reference based approach is that the methods are generally fast, scaling linearly with the number of sequences being clustered.

The second approach has been referred to as distance-based [@Schloss2011] or *de novo* clustering [@NavasMolina2013]. In this method, the distance between sequences is used to cluster sequences into OTUs rather than the distance to a reference database. In contrast to the efficiency of closed-reference clustering, the speed of hierarchical *de novo* clustering methods scale quadratically with the number of unique sequences. The expansion in sequencing throughput combined with sequencing errors inflates the number of unique sequences resulting in the need for large amounts of memory and time to cluster the sequences. If error rates can be reduced through stringent quality control measures, then these problems can be overcome [@Kozich2013]. As an alternative, heuristics have been developed to approximate the clustering of hierarchical methods [@Edgar2010, @Sun2009, @Mah2014]. One critique of *de novo* approaches is that OTU assignments are sensitive to the input order of the sequences [@He2015, @Mah2014]. Whether the differences in assignments is meaningful is unclear; however, the variation in results could represent equally valid clustering of the data. The strength of *de novo* clustering is its independence of references for carrying out the clustering step. After clustering, the classification of each sequence can be used to obtain a consensus classification for the OTU [@Schloss2011]. For this reason, *de novo* clustering has been preferred across the field.

The third approach, open-reference clustering, is a hybrid of the closed-reference and *de novo* approaches [@NavasMolina2013; @Rideout2014]. Open-reference clustering method involves performing closed-reference clustering followed by *de novo* clustering on those sequences that are not sufficiently similar to the reference. In theory, this method should exploit the strengths of both closed-referene and *de novo* clustering; however, the different OTU definitions employed by both approaches poses a possible problem when the methods are combined. An alternative to this approach has been to classify sequences to a bacterial family or genus and then assigned to OTUs within those levels using the average linkage method [@Schloss2011]. For example, all sequences assigned to the *Porphyromonadaceae* would then be assigned to OTUs using the average linkage method using a 3% distance threshold. Those sequences that did not classify to a known family would also be clustered using the average linkage method. An advantage of this approach is that it lends itself nicely to parallelization since each taxonomic group is seen as being independent and can be processed separately. Such an approach would overcome the difficulty of mixing OTU definitions between the closed-reference and *de novo* approaches.

The growth in options for assigning sequences using each of these three broad approaches has been considerable. It has been difficult to objectively assess the quality of OTU assignments. Some have focused on the time and memory required to process a dataset [@Sun2009; @Cai2011; @Rideout2014, @Mah2014]. These are valid parameters to assess when judging a clustering method, but have little to say about the quality of the clustering. Others have attempted to judge the quality of a method by its ability to generate data that parallels classification data [@Cai2011; @Sun2011; @Chen2013; @Edgar2013]. This approach is problematic because bacterial taxonomy often reflects historical biases amongst bacterial systematicists. Furthermore, it is well known that the rates of evolution across lineages are not the same [@Wang2007; @Schloss2010].  Others have assessed the quality of clustering based on their ability to generate the same OTUs generated by other methods [@Schmidt2014; @Rideout2014]. This is problematic because it does not solve the fundamental question of which method is most correct. We recently proposed an approach for evaluating OTU assignments using the distances between pairs of sequences [@Schloss2011]. Those sequences that were similar to each other and found in the same OTU were called true positives while those that were similar and found in different OTUs were called false negatives. Meanwhile, those sequences that were different from each other and found in the same OTU were called false positives and those that were dissimilar and found in different OTUs were called true negatives. Counting the frequency of these different classes allowed us to judge how each method balanced the ratio of true positives and negatives to false positives and negatives using the Matthew's correlation coefficient [MCC; @Matthews1975]. This is an objective approach to assessing the quality of the OTU assignments.

@@

First, we sought to quantify whether the OTU assignments observed for a subset of the data represented the same assignments that were found with the full dataset. The He study found that when they used the open and closed-reference methods the OTUs formed using the subsetted data most closely resembled those of the full dataset. Among the *de novo* methods they observed that the abundance-based greedy clustering (AGC) method generated the most stable OTUs followed by the single linkage (SL), distance-based greedy clustering (DGC), complete linkage (CL), and average linkage (AL) methods. We first sought to assess the calculated the MCC for the OTU assignments generated by each of the clustering methods using 20, 40, 60, and 80% of the sequences relative to the OTU composition formed by the methods using the full dataset (Figure 1A). Similar to the He study, we replicated each method and subsample 30 times because a random number generator is used in some of methods to break ties where pairs of sequences have the same distance between them. Across these sequencing depths, we observed that the stability of the OTUs generated by the SL and CL methods were highly sensitive to sampling effort relative to the OTUs generated by the AL, AGC, and DGC methods (Figure 1A). Our results (Figure 1B) largely confirmed those of Figure 4C in the He study with one notable exception. The He study observed a broad range of MCC values among their AL replicates when analyzing OTUs generated using 60% of the data. This result appeared out of character and was not explained by the authors. They observed a mean MCC value of approximately 0.63 (95% confidence interval between approximately 0.15 and 0.75). In contrast, we observed a mean value of `r round(he_stability[he_stability$method=="an" & he_stability$fraction==0.6, "mean"], 2)` (95% confidence interval between `r round(he_stability[he_stability$method=="an" & he_stability$fraction==0.6, "lci"], 2)` and `r round(he_stability[he_stability$method=="an" & he_stability$fraction==0.6, "uci"], 2)`). This result indicates that the AL assignments were far more stable than indicated in the He study. Regardless, although the assignments are quite stable, it does support the assertion that the OTU assignments observed for the subset of the data do not perfectly match the assignments that were found with the full dataset as they did with the reference-based methods; however, the significance of these differences is unclear.

@@

***OTU assignment methods vary in their accuracy.*** More important than the stability of OTUs is whether sequences are assigned to the correct OTUs. A method can generate highly stable OTUs, but the OTU assignments may be meaningless if they poorly represent the specified cutoff and the actual distance between the sequences. To assess the quality of OTU assignments by the various methods, we made use of the pairwise distance between the unique sequences to count the number of true positives and negatives and the number of false positives and negatives for each method and sampling depth. This enabled us to calculate the average MCC value as a measure of a method's accuracy and its variation as a measure of its precision. We made three important observations. First, each of the *de novo* methods varied in how sensitive their MCC values were to additional sequences (Figure 1C). The SL and CL methods were the most sensitive; however, the accuracy of the OTU assignments did not meaningfully differ when 80 or 100% of the data were assigned to OTUs using the *de novo* methods. Second, the AL method had higher MCC values than the other methods followed by DGC, AGC, CL, open-reference, and closed-reference, and SL (Figure 1D). Third, with the possible exception of the CL method, the MCC values for each of the methods only demonstrated a small amount of variation between runs of the method with a different ordering of the input sequences. This indicates that although there may be variation between executions of the same method, they produce OTU assignments that are equally good. Revisiting the concept of stability, we question the value of obtaining stable OTUs when the full dataset is not optimally assigned to OTUs. Our analysis indicates that the most rigorous method for assigning the Canadian soils sequences to OTUs using a 97% threshold is the AL method.

@@

***Deep sampling of 16S rRNA genes.*** Three factors make the Canadian soil dataset less than desirable to evaluate clustering methods. First, it was one of the earliest 16S rRNA gene sequence datasets published using the 454 FLX platform. Developments in sequencing technology now permit the sequencing of millions of sequences for a study. In addition, because the original Phred quality scores and flowgram data are not available, it was not possible for us to adequately remove sequencing errors [@Schloss2011Reducing]. The large number of sequences that one would expect to remain in the dataset are likely to negatively affect the performance of all of the clustering methods. Second, the dataset used in the He study covered the V9 region of the 16S rRNA gene. For a variety of reasons, this region is not well represented in databases, including the reference database used by the closed and open-reference methods. Of the `r n_v19_total` sequences in the default QIIME database, only `r n_v9_overlap` fully cover the V9 region. In contrast, `r n_v4_overlap` of the sequences fully covered the V4 region. Inadequate coverage of the V9 region would adversely affect the ability of the reference-based methods to assign sequences to OTUs. Third, our previous analysis has shown that the V9 region evolves at a rate much slower than the rest of the gene [@Schloss2010]. With these points in mind, we compared the clustering assignment for each of these methods using a time series experiment that was obtained using mouse feces [@Schloss2012; @Kozich2013]. The MiSeq platform was used to generate `r n_miseq_seqs` sequences from the V4 region of the 16S rRNA gene of `r n_groups` samples. Parallel sequencing of a mock community indicated that the sequencing error rate was approximately 0.02% [@Kozich2013]. Although no dataset is perfect for exhaustively testing these clustering methods, this dataset was useful for demonstrating several points. First, when using 60% of the data the stability relationships amongst the different methods were similar to what we observed using the Canadian soil dataset (Figure 3AB). With the exception for the clusters generated using CL, the methods all performed very well with stabilities greater than `r round(min(miseq_stability[miseq_stability$fraction=="0.6" & miseq_stability$method != "fn","mean"]), 2)`. Second, the MCC values calculated relative to the distances between sequences were generally higher than was observed for the Canadian soil dataset for all of the methods except the CL and SL methods. Surprisingly, the MCC values for the DGC (`r round(miseq_mcc[miseq_mcc$fraction==1.0 & miseq_mcc$method == "dgc","mean"],2)`)  and AGC (`r round(miseq_mcc[miseq_mcc$fraction==1.0 & miseq_mcc$method == "agc","mean"],2)`) methods were comparable to the AL method (`r round(miseq_mcc[miseq_mcc$fraction==1.0 & miseq_mcc$method == "an","mean"],2)`; Figure 3CD). This result suggests that the optimal method is likely to be database-dependent.

@@

***Evalution of an open-source alternative to USEARCH.*** For some datasets the AGC and DGC methods appear to perform as well or better than the hierarchical clustering methods. As originally described in the He study, the AGC and DGC methods utilized the USEARCH program [@Edgar2010], which is not available as open source code and is only available for free to academic users as a 32-bit program. Access for non-academic users and those needing the 64-bit version is available commercially from the developer. An alternative to USEARCH is VSEARCH, which is being developed in parallel to USEARCH as an open-source alternative. One subtle difference between the two programs is that USEARCH employs a heuristic to generate candidate alignments whereas VSEARCH generates the actual global alignments. The VSEARCH developers claim that this difference enhances the sensitivity of VSEARCH relative to USEARCH. Using the two datasets, we determined whether the AGC and DGC methods, as implemented by the two programs, yielded OTU assignments of similar quality. In general the overall trends that we observed with the USEARCH-version of AGC and DGC were also observed with the VSEARCH-verion of the methods (Figure 4). When we compared the two implementations of the AGC and DGC methods, the OTUs generated by the VSEARCH-verion of the methods were as stable or more stable than the USEARCH-version when using 60% of the datasets. In addition, the MCC values for the entire datasets, calculated relative to the distance matrix, were virtually indistinguishable. These results are a strong indication that VSEARCH is a suitable and possibly better replacement for USEARCH for executing the AGC and DGC methods.

***Evalution of Swarm as an alternative de novo clustering algorithm.*** The Swarm algorithm is a recently proposed *de novo* method for assigning sequences to OTUs that identifies clusters of sequences based on the number of differences between each other without applying a distance threshold [@Mah2014]. Swarm was originally validated by comparing the results against the expected clusters formed based on the taxonomic composition of a mock community. Similar to the authors of the He study, the Swarm developers suggest that method are needed that are insensitive to input order. Use of Swarm on the Canadian soil and murine datasets demonstrated that similar to the other *de novo* methods, Swarm's OTU assignments changed as sequences were added (Figures 1A and 3A). When we compared the OTU assignments for both datasets when using all of the sequence data, the variation in MCC values across the 30 randomizations were not meaningfully different (Figures 1D and 3D). Most importantly, when we selected the distance threshold that optimized the MCC value, the quality of the OTU assignments was close to that of the AL assignments when using the Canadian soil dataset and considerably worse than that of the murine dataset (Figures 1D and 3D). Interestingly, the distance thresholds that resulted in the largest MCC values were 0.03 and 0.02 for the Canadian soil and murine datasets, respectively. This suggests that OTU definitions are not consistent across datasets when using the Swarm algorithm. Finally, the Swarm developers indicated that hierarchical *de novo* algorithms were too impractical to use on large MiSeq-generated datasets. Our ability to apply AL to the large mouse dataset and even larger datasets suggests [e.g. @Schubert2015, @Zackular2015] that it is not necessary to sacrifice OTU assignment quality for speed.

@@

***Problems with reference-based clustering in general and as implemented in QIIME.*** The He study and our replication attempt validated that the closed-reference method generated perfectly stable OTUs. This was unsurprising since, by definition, the method represents a one-to-one mapping of reads to a reference and treats the input sequences independently. An important test that was not performed in the He study was to determine whether the clustering was sensitive to the order of the sequences in the database. The default database used in QIIME, which was also used in the He study and by others, contains full-length sequences that are at most 97% similar to each other. We randomized the order of the reference sequences 30 times and used them to carry out the closed-reference method with the full murine dataset, which contained `r n_unique_miseq_seqs` unique sequences. Surprisingly, we observed that the number of OTUs generated was not the same in each of the randomizations. On average there were `r round(rand_ref_u["mean", 1])` sequences that mapped to a reference OTU per randomization (range from `r rand_ref_u["min", 1]` to `r rand_ref_u["max", 1]`). The original ordering of the reference resulted in `r rand_ref_u["original", 1]` sequences being mapped, less than the minimum observed number of mapped sequences when the references were randomized. This surprising result was likely due to the performance of the USEARCH heuristic. To test this further, we substituted VSEARCH for USEARCH in the closed-reference method. When we used VSEARCH the original ordering of the reference sequences and all randomizations were able to map `r rand_ref_v["original", 1]` sequences to reference OTUs. When we calculated the true distance between each of the murine sequences and the references, we were able to map `r nrow(miseq_closed_ref)` of the murine sequences to the reference sequences when using a 97% similarity threshold without the use of a heuristic. This result indicates that the closed reference approach, whether using USEARCH or VSEARCH, does not exhaustively or accurately map reads to the closest reference. To quantify this further, we calculated the MCC for the USEARCH and VSEARCH assignments relative to the assignments using the non-heuristic approach. Using USEARCH the average MCC was `r mean(uclosed$mcc)` (range: `r min(uclosed$mcc)` to `r max(uclosed$mcc)`) and using VSEARCH the average MCC was `r mean(vclosed$mcc)` (range: `r min(vclosed$mcc)` to `r max(vclosed$mcc)`). The two methods had similar sensitivities (USEARCH: `r mean(uclosed$sensitivity)` and VSEARCH: `r mean(vclosed$sensitivity)`), but the USEARCH specificity (`r mean(uclosed$specificity)`) was considerably higher than VSEARCH (`r mean(vclosed$specificity)`). Overall, these results indicate that although heuristic approaches may be fast, relative to non-heuristic approaches, they do a poor job of mapping reads to the correct reference sequence.

@@

***Analysis of reference database.***
We utilized the 97% OTUs greengenes reference sequence and taxonomy data (v.13.8) that accompanies the QIIME installation. Because the greengenes reference alignment does a poor job of representing the secondary structure of the 16S rRNA gene [@Schloss2010], we realigned the fasta sequences to a SILVA reference alignment to identify the V4 region of the sequences.

>>>>

**Methods.** Our analysis implemented five *de novo* clustering algorithms including the single linkage, complete linkage, average linkage, abundance-based greedy clustering, distance-based greedy clustering, and the open and closed-reference methods. Using two previously published datasets we used the Matthew’s Correlation Coefficient (MCC) to assess the stability and quality of OTU assignments.

@@

**Discussion.** Our analysis calls into question the quality and stability of OTU assignments generated by the open and closed-reference methods as implemented in current version of QIIME. This study demonstrates that *de novo* methods are the most rigorous and that the quality of clustering assignments needs to be assessed for multiple methods to identify the optimal clustering method for a particular dataset.

@@

The ability to affordably generate millions of 16S rRNA gene sequences has allowed microbial ecologists to thoroughly characterize the microbial community composition of hundreds of samples. To simplify the complexity of these large datasets, it is helpful to cluster sequences into meaningful bins. These bins, commonly known as operational taxonomic units (OTUs), are  used to compare the biodiversity contained within and between different samples [@Schloss2011]. Such comparisons have enabled researchers to characterize the microbiota associated with the human body [e.g. @Huttenhower2012], soil [e.g. @Shade2013], aquatic ecosystems [e.g. @Gilbert2011], and numerous other environments. Within the field of microbial ecology, a convention has emerged where sequences are clustered into OTUs using a threshold of 97% similarity or a distance of 3%. One advantage of the OTU-based approach is that the definition of the bins is operational and can be changed to suit the needs of the particular project. However, with the dissemination of clustering methods within software such as mothur [@Schloss2009], QIIME [@Caporaso2010], and other tools [@Edgar2010, @Sun2009, @Mah2014, @Edgar2013, @Cai2011], it is important to understand how different clustering methods implement this conventional OTU threshold. Furthermore, it is necessary to understand how the selected method affects the precision and accuracy of assigning sequences to OTUs. Broadly speaking, three approaches have been developed to assign sequences to OTUs.

The first approach has been referred to as phylotyping [@Schloss2011] or closed reference clustering [@NavasMolina2013]. This approach involves comparing sequences to a curated database and then clustering sequences into the same OTU that are similar to the same reference sequence. Reference-based clustering methods suffer when the reference does not adequately reflect the biodiversity of the community. If a large fraction of sequences are novel, then they cannot be assigned to an OTU. In addition, the reference sequences are selected because they are less than 97% similar to each other over the full length of the gene; however, it is known that the commonly used variable regions within the 16S rRNA gene do not evolve at the same rate as the full-length gene [@Schloss2011]. Thus, a sequence representing a fragment of the gene may be more than 97% similar to multiple reference sequences. Therefore, defining OTUs in the closed-reference approach is problematic because two sequences might be 97% similar to the same reference sequence, but they may only be 94% similar to each other. Alternatively, a seuquence may be equally similar to two or more reference sequences. A subtle alternative to this approach is to use a classifier to assign a taxonomy to each sequence so that sequences can be clustered at a desired level within the Linnean taxonomic hierarchy [@Schloss2011]. The strength of the reference based approach is that the methods are generally fast, scaling linearly with the number of sequences being clustered.

The second approach has been referred to as distance-based [@Schloss2011] or *de novo* clustering [@NavasMolina2013]. In this approach, the distance between sequences is used to cluster sequences into OTUs rather than the distance to a reference database. In contrast to the efficiency of closed-reference clustering, the speed of hierarchical *de novo* clustering methods scale quadratically with the number of unique sequences. The expansion in sequencing throughput combined with sequencing errors inflates the number of unique sequences resulting in the need for large amounts of memory and time to cluster the sequences. If error rates can be reduced through stringent quality control measures, then these problems can be overcome [@Kozich2013]. As an alternative, heuristics have been developed to approximate the clustering of hierarchical methods [@Edgar2010, @Sun2009, @Mah2014]. One critique of *de novo* approaches is that OTU assignments are sensitive to the input order of the sequences [@He2015, @Mah2014]. Whether the differences in assignments is meaningful is unclear; however, the variation in results could represent equally valid clustering of the data. The strength of *de novo* clustering is its independence of references for carrying out the clustering step. After clustering, the classification of each sequence can be used to obtain a consensus classification for the OTU [@Schloss2011]. For this reason, *de novo* clustering has been preferred across the field.

The third approach, open-reference clustering, is a hybrid of the closed-reference and *de novo* approaches [@NavasMolina2013; @Rideout2014]. Open-reference clustering involves performing closed-reference clustering followed by *de novo* clustering on those sequences that are not sufficiently similar to the reference. In theory, this method should exploit the strengths of both closed-reference and *de novo* clustering; however, the different OTU definitions employed by closed-reference and *de novo* clustering pose a possible problem when the methods are combined. An alternative to this approach has been to classify sequences to a bacterial family or genus and then assigned to OTUs within those levels using the average linkage method [@Schloss2011]. For example, all sequences assigned to the *Porphyromonadaceae* would then be assigned to OTUs using the average linkage method using a 3% distance threshold. Those sequences that did not classify to a known family would also be clustered using the average linkage method. An advantage of this approach is that it lends itself nicely to parallelization since each taxonomic group is seen as being independent and can be processed separately. Such an approach would overcome the difficulty of mixing OTU definitions between the closed-reference and *de novo* approaches.

The growth in options for assigning sequences using each of these three broad approaches has been considerable. It has been difficult to objectively assess the quality of OTU assignments. Some have focused on the time and memory required to process a dataset [@Sun2009; @Cai2011; @Rideout2014, @Mah2014]. These are valid parameters to assess when judging a clustering method, but have little to say about the quality of the OTU assignments. Others have attempted to judge the quality of a method by its ability to generate data that parallels classification data [@Cai2011; @Sun2011; @Chen2013; @Edgar2013]. This approach is problematic because bacterial taxonomy often reflects historical biases amongst bacterial systematicists. Furthermore, it is well known that the rates of evolution across lineages are not the same [@Wang2007; @Schloss2010].  Others have assessed the quality of clustering based on their ability to generate the same OTUs generated by other methods [@Schmidt2014; @Rideout2014]. This is problematic because it does not solve the fundamental question of which method is most correct. We recently proposed an approach for evaluating OTU assignments using the distances between pairs of sequences [@Schloss2011]. Those sequences that were similar to each other and found in the same OTU were called true positives while those that were similar and found in different OTUs were called false negatives. Meanwhile, those sequences that were different from each other and found in the same OTU were called false positives and those that were dissimilar and found in different OTUs were called true negatives. Counting the frequency of these different classes allowed us to judge how each method balanced the ratio of true positives and negatives to false positives and negatives using the Matthew's correlation coefficient [MCC; @Matthews1975]. This is an objective approach to assessing the quality of the OTU assignments.

@@

First, we sought to quantify whether the OTU assignments observed for a subset of the data represented the same assignments that were found with the full dataset. The He study found that when they used the open and closed-reference methods the OTUs formed using the subsetted data most closely resembled those of the full dataset. Among the *de novo* methods they observed that the abundance-based greedy clustering (AGC) method generated the most stable OTUs followed by the single linkage (SL), distance-based greedy clustering (DGC), complete linkage (CL), and average linkage (AL) methods. We first sought to assess the calculated the MCC for the OTU assignments generated by each of the clustering methods using 20, 40, 60, and 80% of the sequences relative to the OTU composition formed by the methods using the full dataset (Figure 1A). Similar to the He study, we replicated each method and subsampled to the desired fraction of the dataset 30 times. Multiple subsamplings was necessary because a random number generator is used in some of the methods to break ties where pairs of sequences have the same distance between them. Across these sequencing depths, we observed that the stability of the OTUs generated by the SL and CL methods were highly sensitive to sampling effort relative to the OTUs generated by the AL, AGC, and DGC methods (Figure 1A). Our results (Figure 1B) largely confirmed those of Figure 4C in the He study with one notable exception. The He study observed a broad range of MCC values among their AL replicates when analyzing OTUs generated using 60% of the data. This result appeared out of character and was not explained by the authors. They observed a mean MCC value of approximately 0.63 (95% confidence interval between approximately 0.15 and 0.75). In contrast, we observed a mean value of `r round(he_stability[he_stability$method=="an" & he_stability$fraction==0.6, "mean"], 2)` (95% confidence interval between `r round(he_stability[he_stability$method=="an" & he_stability$fraction==0.6, "lci"], 2)` and `r round(he_stability[he_stability$method=="an" & he_stability$fraction==0.6, "uci"], 2)`). This result indicates that the AL assignments were far more stable than indicated in the He study. Regardless, although the assignments are quite stable, it does support the assertion that the OTU assignments observed for the subset of the data do not perfectly match the assignments that were found with the full dataset as they did with the reference-based methods; however, the significance of these differences is unclear.

@@

***OTU assignment methods vary in their accuracy.*** More important than the stability of OTUs is whether sequences are assigned to the correct OTUs. A method can generate highly stable OTUs, but the OTU assignments may be meaningless if they poorly represent the specified cutoff and the actual distance between the sequences. To assess the quality of OTU assignments by the various methods, we made use of the pairwise distance between the unique sequences to count the number of true positives and negatives and the number of false positives and negatives for each method and sampling depth. This enabled us to calculate the average MCC value as a measure of a method's accuracy and its variation as a measure of its precision. We made three important observations. First, each of the *de novo* methods varied in how sensitive their MCC values were to additional sequences (Figure 1C). The SL and CL methods were the most sensitive; however, the accuracy of the OTU assignments did not meaningfully differ when 80 or 100% of the data were assigned to OTUs using the *de novo* methods. Second, the AL method had higher MCC values than the other methods followed by DGC, AGC, CL, open-reference, and closed-reference, and SL (Figure 1D). Third, with the possible exception of the CL method, the MCC values for each of the methods only demonstrated a small amount of variation between runs of the method with a different ordering of the input sequences. This indicates that although there may be variation between executions of the same method, they produce OTU assignments that are equally good. Revisiting the concept of stability, we question the value of obtaining stable OTUs when the full dataset is not optimally assigned to OTUs. Our analysis indicates that the most rigorous method for assigning the Canadian soils sequences to OTUs using a 97% threshold was the AL method.

@@

***Deep sampling of 16S rRNA genes.*** Three factors make the Canadian soil dataset less than desirable to evaluate clustering methods. First, it was one of the earliest 16S rRNA gene sequence datasets published using the 454 FLX platform. Developments in sequencing technology now permit the sequencing of millions of sequences for a study. In addition, because the original Phred quality scores and flowgram data are not available, it was not possible for us to adequately remove sequencing errors [@Schloss2011Reducing]. The large number of sequences that one would expect to remain in the dataset are likely to negatively affect the performance of all of the clustering methods. Second, the dataset used in the He study covered the V9 region of the 16S rRNA gene. For a variety of reasons, this region is not well represented in databases, including the reference database used by the closed and open-reference methods. Of the `r n_v19_total` sequences in the default QIIME database, only `r n_v9_overlap` fully cover the V9 region. In contrast, `r n_v4_overlap` of the sequences fully covered the V4 region. Inadequate coverage of the V9 region would adversely affect the ability of the reference-based methods to assign sequences to OTUs. Third, our previous analysis has shown that the V9 region evolves at a rate much slower than the rest of the gene [@Schloss2010]. With these points in mind, we compared the clustering assignment for each of these methods using a time series experiment that was obtained using mouse feces [@Schloss2012; @Kozich2013]. The MiSeq platform was used to generate `r n_miseq_seqs` sequences from the V4 region of the 16S rRNA gene of `r n_groups` samples. Parallel sequencing of a mock community indicated that the sequencing error rate was approximately 0.02% [@Kozich2013]. Although no dataset is perfect for exhaustively testing these clustering methods, this dataset was useful for demonstrating several points. First, when using 60% of the data, the stability relationships amongst the different methods were similar to what we observed using the Canadian soil dataset (Figure 3AB). With the exception for the clusters generated using CL, the methods all performed very well with stabilities greater than `r round(min(miseq_stability[miseq_stability$fraction=="0.6" & miseq_stability$method != "fn","mean"]), 2)`. Second, the MCC values calculated relative to the distances between sequences were generally higher than was observed for the Canadian soil dataset for all of the methods except the CL and SL methods. Surprisingly, the MCC values for the DGC (`r round(miseq_mcc[miseq_mcc$fraction==1.0 & miseq_mcc$method == "dgc","mean"],2)`)  and AGC (`r round(miseq_mcc[miseq_mcc$fraction==1.0 & miseq_mcc$method == "agc","mean"],2)`) methods were comparable to the AL method (`r round(miseq_mcc[miseq_mcc$fraction==1.0 & miseq_mcc$method == "an","mean"],2)`; Figure 3CD). This result suggests that the optimal method is likely to be database-dependent.

@@

***Evalution of Swarm as an alternative de novo clustering algorithm.*** The Swarm algorithm is a recently proposed *de novo* method for assigning sequences to OTUs that identifies clusters of sequences based on the number of differences between each other without applying a distance threshold [@Mah2014]. Swarm was originally validated by comparing the results against the expected clusters formed based on the taxonomic composition of a mock community. Similar to the authors of the He study, the Swarm developers suggest that method are needed that are insensitive to input order. Use of Swarm on the Canadian soil and murine datasets demonstrated that similar to the other *de novo* methods, Swarm's OTU assignments changed as sequences were added (Figures 1A and 3A). When we compared the OTU assignments for both datasets when using all of the sequence data, the variation in MCC values across the 30 randomizations were not meaningfully different (Figures 1D and 3D). Most importantly, when we selected the distance threshold that optimized the MCC value, the quality of the OTU assignments was close to that of the AL assignments when using the Canadian soil dataset and considerably worse than that of the murine dataset (Figures 1D and 3D). Interestingly, the distance thresholds that resulted in the largest MCC values were 3 and 2% for the Canadian soil and murine datasets, respectively. This suggests that OTU definitions are not consistent across datasets when using the Swarm algorithm. Finally, the Swarm developers indicated that hierarchical *de novo* algorithms were too impractical to use on large MiSeq-generated datasets. Our ability to apply AL to the large mouse dataset and even larger datasets suggests [e.g. @Schubert2015, @Zackular2015] that it is not necessary to sacrifice OTU assignment quality for speed.

***Evalution of an open-source alternative to USEARCH.*** For some datasets the AGC and DGC methods appear to perform as well or better than the hierarchical clustering methods. As originally described in the He study, the AGC and DGC methods utilized the USEARCH program and the DGC method is used for clustering in UPARSE [@Edgar2010, @Edgar2013]. The source code for USEARCH is not publicly available and only the 32-bit executables are available for free to academic users. Access for non-academic users and those needing the 64-bit version is available commercially from the developer. An alternative to USEARCH is VSEARCH, which is being developed in parallel to USEARCH as an open-source alternative. One subtle difference between the two programs is that USEARCH employs a heuristic to generate candidate alignments whereas VSEARCH generates the actual global alignments. The VSEARCH developers claim that this difference enhances the sensitivity of VSEARCH relative to USEARCH. Using the two datasets, we determined whether the AGC and DGC methods, as implemented by the two programs, yielded OTU assignments of similar quality. In general the overall trends that we observed with the USEARCH-version of AGC and DGC were also observed with the VSEARCH-verion of the methods (Figure 4). When we compared the two implementations of the AGC and DGC methods, the OTUs generated by the VSEARCH-verion of the methods were as stable or more stable than the USEARCH-version when using 60% of the datasets. In addition, the MCC values for the entire datasets, calculated relative to the distance matrix, were virtually indistinguishable. These results are a strong indication that VSEARCH is a suitable and possibly better option for executing the AGC and DGC methods.

@@

***Problems with reference-based clustering in general and as implemented in QIIME.*** The He study and our replication attempt validated that the closed-reference method generated perfectly stable OTUs. This was unsurprising since, by definition, the method is designed to return one-to-one mapping of reads to a reference. Furthermore, because it treats the input sequences independently the input order or use of a random number generator is not an issue. An important test that was not performed in the He study was to determine whether the clustering was sensitive to the order of the sequences in the database. The default database used in QIIME, which was also used in the He study, contains full-length sequences that are at most 97% similar to each other. We randomized the order of the reference sequences 30 times and used them to carry out the closed-reference method with the full murine dataset, which contained `r n_unique_miseq_seqs` unique sequences. Surprisingly, we observed that the number of OTUs generated was not the same in each of the randomizations. On average there were `r round(rand_ref_u["mean", 1])` sequences that mapped to a reference OTU per randomization (range from `r rand_ref_u["min", 1]` to `r rand_ref_u["max", 1]`). The original ordering of the reference resulted in `r rand_ref_u["original", 1]` sequences being mapped, less than the minimum observed number of mapped sequences when the references were randomized. This surprising result was likely due to the performance of the USEARCH heuristic. To test this further, we substituted VSEARCH for USEARCH in the closed-reference method. When we used VSEARCH the original ordering of the reference sequences and all randomizations were able to map `r rand_ref_v["original", 1]` sequences to reference OTUs. When we calculated the true distance between each of the murine sequences and the references, we were able to map `r nrow(miseq_closed_ref)` of the murine sequences to the reference sequences when using a 97% similarity threshold without the use of a heuristic. This result indicates that the closed reference approach, whether using USEARCH or VSEARCH, does not exhaustively or accurately map reads to the closest reference. To quantify this further, we calculated the MCC for the USEARCH and VSEARCH assignments relative to the assignments using the non-heuristic approach. Using USEARCH the average MCC was `r mean(uclosed$mcc)` (range: `r min(uclosed$mcc)` to `r max(uclosed$mcc)`) and using VSEARCH the average MCC was `r mean(vclosed$mcc)` (range: `r min(vclosed$mcc)` to `r max(vclosed$mcc)`). The two methods had similar sensitivities (USEARCH: `r mean(uclosed$sensitivity)` and VSEARCH: `r mean(vclosed$sensitivity)`), but the USEARCH specificity (`r mean(uclosed$specificity)`) was considerably higher than VSEARCH (`r mean(vclosed$specificity)`). Overall, these results indicate that although heuristic approaches may be fast, they do a poor job of mapping reads to the correct reference sequence relative to non-heuristic approaches.

@@

***Analysis of reference database.***
We utilized the 97% OTUs greengenes reference sequence and taxonomy data (v.13.8) that accompanies the QIIME installation. Because the greengenes reference alignment does a poor job of representing the secondary structure of the 16S rRNA gene [@Schloss2010], we realigned the FASTA sequences to a SILVA reference alignment to identify the V4 region of the sequences.
